---
title: "TEDS_SH_Age_Twin_Markdown"
author: "Filip Kaleta"
date: "2023-02-23"
output: html_document
---

```{r}


#variables:


#Sex at 1st contact: sex <- teds$sex1

#Gender Identity at 26: gender_id <- teds$zmhgender1

#Self-Harm past 12m 26: sh_26 <- teds$zmhslfhm021

```

#Load packages

```{r include=FALSE}
library(psych)
library(ggplot2)
library(scales)
library(stringr)
library(dplyr)
library(forcats)
library(tidyverse)
library(haven)
library(labelled)
library(car)

require(OpenMx)
```

#Load data

```{r}
load("TEDS_SH_Age.rdata")

teds <- df
```

#Examine data

```{r}
dim(teds)
head(teds)
tail(teds)
```

#Objects of Key Variables

zmhslfhm022

```{r include=FALSE}

as.factor(teds$sex1)
sex1 <- teds$sex1
sex2 <- teds$sex2

as_factor(teds$zmhslfhm021)
as_factor(teds$zmhslfhm022)


```

#Recode SH at 26

```{r include=FALSE}

#Self-harm at 21 will also be dichotomised for the genetic analysis. The “No” response to the question on self-harm will be coded as 0, and all the other responses will be coded as 1 (“Yes, once or twice”, “Yes, 3-5 times”, “Yes, 6-10 times” and “Yes, more than 10 times”). 

val_labels(teds$zmhslfhm021)

#Dichotomising SH 26 TWIN 1 

teds$sh_26_2_1[teds$zmhslfhm021 == 0] <- 0
teds$sh_26_2_1[teds$zmhslfhm021 == 1] <- 1
teds$sh_26_2_1[teds$zmhslfhm021 == 2] <- 1
teds$sh_26_2_1[teds$zmhslfhm021 == 3] <- 1
teds$sh_26_2_1[teds$zmhslfhm021 == 4] <- 1
teds$sh_26_2_1[teds$zmhslfhm021 == 5] <- 1

#Dichotomising SH 26 TWIN 2

teds$sh_26_2_2[teds$zmhslfhm022 == 0] <- 0
teds$sh_26_2_2[teds$zmhslfhm022 == 1] <- 1
teds$sh_26_2_2[teds$zmhslfhm022 == 2] <- 1
teds$sh_26_2_2[teds$zmhslfhm022 == 3] <- 1
teds$sh_26_2_2[teds$zmhslfhm022 == 4] <- 1
teds$sh_26_2_2[teds$zmhslfhm022 == 5] <- 1


sh_26_2_1 <- teds$sh_26_2_1
sh_26_2_2 <- teds$sh_26_2_2

```

#Descriptives

```{r}

sum(!is.na(sh_26_2_1))

table(sh_26_2_1)
prop.table(table(sh_26_2_1))

table(sh_26_2_2)
prop.table(table(sh_26_2_2))


```

```{r}

teds <- teds[teds$random==1 & teds$exclude1==0,]

#Make the integer variables ordered factors
teds$sh_26_2_1	<-mxFactor(teds$sh_26_2_1, levels=c(0:1) )
teds$sh_26_2_2 <- mxFactor(teds$sh_26_2_2, levels=c(0:1) )


nv 	<- 1		# number of variables per twin
ntv 	<- nv*2	# number of variables per pair
nth	<- 1		# number of max thresholds

selVars	<- c('sh_26_2_1' , 'sh_26_2_2')

# Select Data for Analysis
mzData <- subset(teds, zygos==1, selVars)
dzData <- subset(teds, zygos==2, selVars)
describe(mzData)
describe(dzData)

# get CT for Ordinal variable
table(mzData$sh_26_2_1, mzData$sh_26_2_2)
table(dzData$sh_26_2_1, dzData$sh_26_2_2)
```

\# 1) Specify Saturated Model (Polychoric correlations)

```{r}

# Matrix & Algebra for expected means (SND), Cor, Thresholds, Age effect on Th

MeanL		<-mxMatrix( type="Zero", nrow=1, ncol=ntv, name="M" )

Tmz		<-mxMatrix( type="Full", nrow=nth, ncol=ntv, free=T, values=c(.91, .91), 

					labels=c("Tmz11","Tmz12"), name="ThMZ")

CorMZ		<-mxMatrix( type="Stand", nrow=ntv, ncol=ntv, free=T, values=.6, 
					lbound=-.99, ubound=.99, name="expCorMZ") 

Tdz		<-mxMatrix( type="Full", nrow=nth, ncol=ntv, free=T, values=c(.91, .91), 
		
					labels=c("Tdz11","Tdz12"), name="ThDZ")

CorDZ		<-mxMatrix( type="Stand", nrow=ntv, ncol=ntv, free=T, values=.3, 
					lbound=-.99, ubound=.99, name="expCorDZ") 

# Data objects for Multiple Groups

dataMZ   	<- mxData( observed=mzData, type="raw" )
dataDZ   	<- mxData( observed=dzData, type="raw" )

# Objective objects for Multiple Groups

objMZ    	<- mxExpectationNormal(covariance="expCorMZ", means="M", dimnames=selVars, thresholds="ThMZ" )

objDZ    	<- mxExpectationNormal(covariance="expCorDZ", means="M", dimnames=selVars, thresholds="ThDZ" )

fitFunction <- mxFitFunctionML()

# Combine Groups
modelMZ	<- mxModel(MeanL, Tmz, CorMZ, dataMZ, objMZ, fitFunction, name="MZ" )
modelDZ	<- mxModel(MeanL, Tdz, CorDZ, dataDZ, objDZ, fitFunction, name="DZ" )
minus2ll	<- mxAlgebra(expression=MZ.objective + DZ.objective, name="m2LL" )
obj		<- mxFitFunctionAlgebra( "m2LL" )
Conf	<- mxCI(c('MZ.expCorMZ[2,1]', 'DZ.expCorDZ[2,1]'))
SatModel 	<- mxModel( "Sat", modelMZ, modelDZ, minus2ll, obj, Conf)
```

\# 1) RUN Saturated Model

```{r}
#SatFit	<- mxRun(SatModel, intervals=TRUE)

SatFit		<- mxTryHardOrdinal(SatModel, intervals=T)
(SatSumm		<- summary(SatFit))


# Different ways of getting output:
round(SatFit@output$estimate,4)
SatFit$MZ.expCorMZ
SatFit$DZ.expCorDZ

# Generate output using OpenMx functions
mxEval(MZ.expCorMZ, SatFit)
mxEval(DZ.expCorDZ, SatFit)
```

\# 2) Specify and Run Sub1Model: equating Thresholds across Twin 1 and Twin 2 \# within MZ and DZ group

```{r}
Sub1Model	<- mxModel(SatModel, name="sub1")
Sub1Model	<- omxSetParameters(Sub1Model, labels=c("Tmz11","Tmz12"), 
							 newlabels=c("Tmz11","Tmz11"), 
							 'strict'=FALSE, free=TRUE, values=.91, lbound=-3, ubound=3)
Sub1Model	<- omxSetParameters(Sub1Model, labels=c("Tdz11","Tdz12"), 
							 newlabels=c("Tdz11","Tdz11"), 
							 'strict'=FALSE, free=TRUE, values=.91, lbound=-3, ubound=3)
Sub1Fit	<- mxTryHardOrdinal(Sub1Model, intervals=TRUE)
(Sub1Summ	<- summary(Sub1Fit))

# Generate Sub1Model Output
round(Sub1Fit@output$estimate,4)
mxEval(MZ.expCorMZ, Sub1Fit)
mxEval(DZ.expCorDZ, Sub1Fit)
mxCompare(SatFit,Sub1Fit)
```

\# 3) Specify and Run Sub2Model: equating Thresh across Twins and zygosity groups

```{r}
Sub2Model	<- mxModel(SatModel, name="sub2")
Sub2Model	<- omxSetParameters(Sub2Model, labels=c("Tmz11","Tmz12"), 
							 newlabels=c("Tmz11","Tmz11"), 
							 'strict'=FALSE, free=TRUE, values=.91, lbound=-3, ubound=3)
Sub2Model	<- omxSetParameters(Sub2Model, labels=c("Tdz11","Tdz12"), 
							 newlabels=c("Tmz11","Tmz11"), 
							 'strict'=FALSE, free=TRUE, values=.91, lbound=-3, ubound=3)
Sub2Fit	<- mxTryHardOrdinal(Sub2Model, intervals=TRUE)
(Sub2Summ	<- summary(Sub2Fit))

# Generate Sub2Model Output
round(Sub2Fit@output$estimate,4)
mxEval(MZ.expCorMZ, Sub2Fit)
mxEval(DZ.expCorDZ, Sub2Fit)
mxCompare(SatFit,Sub2Fit)
```

#4) Specify ACE Model, with ONE overall set of Thresholds (withOUT Age effects on thresholds)

```{r}

# Matrix & Algebra for expected means (SND), Thresholds and correlation
MeanL		<-mxMatrix( type="Zero", nrow=1, ncol=ntv, name="M")

add		<- mxMatrix(type="Lower", nrow=nth, ncol=nth, free=F, values=1, name="Low") 
TH		<- mxMatrix(type="Full", nrow=nth, ncol=ntv, free=T, values= .91, lbound= -3, ubound= 3, labels="T1", name="Th")

# Matrices declared to store a, c, and e Path Coefficients
pathA	<- mxMatrix( type="Lower", nrow=nv, ncol=nv, free=T, values=.8, label="a11", name="a" ) 
pathC	<- mxMatrix( type="Lower", nrow=nv, ncol=nv, free=T, values=.3, label="c11", name="c" )
pathE	<- mxMatrix( type="Lower", nrow=nv, ncol=nv, free=T, values=.6, label="e11", name="e" )
	
# Matrices generated to hold A, C, and E components and total Variance
covA	<- mxAlgebra( expression=a %*% t(a), name="A" )
covC	<- mxAlgebra( expression=c %*% t(c), name="C" ) 
covE	<- mxAlgebra( expression=e %*% t(e), name="E" )
covP	<- mxAlgebra( expression=A+C+E, name="V" )

# Algebra for expected Variance/Covariance Matrices in MZ & DZ twins
covMZ	<- mxAlgebra( expression= rbind( cbind(A+C+E , A+C),
                                       cbind(A+C   , A+C+E)), name="expCovMZ" )
covDZ	<- mxAlgebra( expression= rbind( cbind(A+C+E     , 0.5%x%A+C),
                                       cbind(0.5%x%A+C , A+C+E)), name="expCovDZ" )

# Constraint on total variance of Ordinal variables (A+C+E=1)
mUnv	<- mxMatrix( type="Unit", nrow=nv, ncol=1, name="Unv" )
varL	<- mxConstraint( expression=diag2vec(V)==Unv, name="VarL" )

# Data objects for Multiple Groups
dataMZ   <- mxData( observed=mzData, type="raw" )
dataDZ   <- mxData( observed=dzData, type="raw" )

# Objective objects for Multiple Groups
objMZ	<- mxExpectationNormal( covariance="expCovMZ", means="M", dimnames=selVars, thresholds="Th" )
objDZ	<- mxExpectationNormal( covariance="expCovDZ", means="M", dimnames=selVars, thresholds="Th" )
fitFunction <- mxFitFunctionML()

# Combine Groups
pars		<- list(MeanL, TH, add, pathA, pathC, pathE, covA, covC, covE, covP, mUnv, varL )
modelMZ	<- mxModel(pars, covMZ, dataMZ, objMZ, fitFunction, name="MZ" )
modelDZ	<- mxModel(pars, covDZ, dataDZ, objDZ, fitFunction, name="DZ" )
minus2ll	<- mxAlgebra( expression=MZ.objective + DZ.objective, name="m2LL" )
obj		<- mxFitFunctionAlgebra( "m2LL" )
Conf <- mxCI(c('MZ.A[1,1]', 'MZ.C[1,1]', 'MZ.E[1,1]'))
AceModel 	<- mxModel( "ACE", modelMZ, modelDZ, minus2ll, obj, Conf)
```

\# 4) RUN AceModel

```{r}
AceFit   <- mxTryHardOrdinal(AceModel, intervals=TRUE)
(AceSumm  <- summary(AceFit, verbose = T))
round(AceFit@output$estimate,4)

mxEval(MZ.A, AceFit)
mxEval(MZ.C, AceFit)
mxEval(MZ.E, AceFit)

mxCompare (SatFit,AceFit)
```

# Specify ACDE model to run an ADE model

```{r}

# Matrix & Algebra for expected means (SND), Thresholds and correlation
MeanL		<-mxMatrix( type="Zero", nrow=1, ncol=ntv, name="M" )


TH		<-mxMatrix( type="Full", nrow=nth, ncol=ntv, free=T, values=c(.82), lbound=c(-3), ubound=3, labels=c("T1"), name="Th") 

# Matrices declared to store a, c, and e Path Coefficients
pathA	<- mxMatrix( type="Lower", nrow=nv, ncol=nv, free=T, values=.8, label="a11", name="a" ) 
pathC	<- mxMatrix( type="Lower", nrow=nv, ncol=nv, free=T, values=.3, label="c11", name="c")
pathD <- mxMatrix( type="Full", nrow=nv, ncol=nv, free=T, values=.3, label="d11", name="d")
pathE	<- mxMatrix( type="Lower", nrow=nv, ncol=nv, free=T, values=.6, label="e11", name="e" )
	
# Matrices generated to hold A, C, and E components and total Variance
covA	<- mxAlgebra( expression=a %*% t(a), name="A" )
covC	<- mxAlgebra( expression=c %*% t(c), name="C" ) 
covD	<- mxAlgebra( expression=d %*% t(d), name="D" ) 
covE	<- mxAlgebra( expression=e %*% t(e), name="E" )
covP	<- mxAlgebra( expression=A+C+E, name="V" )

# Algebra for expected Variance/Covariance Matrices in MZ & DZ twins
covMZ	<- mxAlgebra( expression= rbind( cbind(A+C+E , A+C+D),
                                       cbind(A+C+D , A+C+E)), name="expCovMZ" )
covDZ	<- mxAlgebra( expression= rbind( cbind(A+C+E     , 0.5%x%A+C+0.25%x%D),
                                       cbind(0.5%x%A+C+0.25%x%D, A+C+E)), name="expCovDZ" )

# Constraint on total variance of Ordinal variables (A+C+E=1)
mUnv	<- mxMatrix( type="Unit", nrow=nv, ncol=1, name="Unv" )
varL	<- mxConstraint( expression=diag2vec(V)==Unv, name="VarL" )

# Data objects for Multiple Groups
dataMZ   <- mxData( observed=mzData, type="raw" )
dataDZ   <- mxData( observed=dzData, type="raw" )

# Objective objects for Multiple Groups
objMZ	<- mxExpectationNormal( covariance="expCovMZ", means="M", dimnames=selVars, thresholds="Th" )
objDZ	<- mxExpectationNormal( covariance="expCovDZ", means="M", dimnames=selVars, thresholds="Th" )
fitFunction <- mxFitFunctionML()

# Combine Groups
pars		<- list(MeanL, TH, pathA, pathC, pathD, pathE, covA, covC, covD, covE, covP, mUnv, varL )
modelMZ	<- mxModel(pars, covMZ, dataMZ, objMZ, fitFunction, name="MZ" )
modelDZ	<- mxModel(pars, covDZ, dataDZ, objDZ, fitFunction, name="DZ" )
minus2ll	<- mxAlgebra( expression=MZ.objective + DZ.objective, name="m2LL" )
obj		<- mxFitFunctionAlgebra( "m2LL" )
Conf <- mxCI(c('MZ.A[1,1]', 'MZ.C[1,1]', 'MZ.E[1,1]'))
AdceModel 	<- mxModel("ADCE", modelMZ, modelDZ, minus2ll, obj, Conf)
```

#Run ACDE model first
```{r}
fitADCE <- mxTryHardOrdinal(AdceModel, interval=F)
(sumADCE <- summary(fitADCE))

mxEval(MZ.A, fitADCE)
mxEval(MZ.C, fitADCE)
mxEval(MZ.D, fitADCE)
mxEval(MZ.E, fitADCE)
```
#Run ADE Model
```{r}
modelADE <- mxModel(fitADCE, name = "ADEmodel")
modelADE <- omxSetParameters(modelADE, labels="c11", free=F, values = 0)
AdeFit <- mxTryHardOrdinal(modelADE, intervals=F)
(summary(AdeFit))

round(AdeFit@output$estimate,4)

mxEval(MZ.A, AdeFit)
mxEval(MZ.D, AdeFit)
mxEval(MZ.E, AdeFit)

mxCompare (AceFit, AdeFit)
```